---
title: "Machine learning and statistical inference on persistence diagrams with TDAInference"
author: "Shael Brown"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: Robinson_Turner
  title: 'Hypothesis testing for topological data analysis'
  author:
  - family: Robinson
    given: Andrew
  - family: Turner
    given: Katharine
  container-title: Journal of Applied and Computational Topology
  volume: 1
  URL: 'https://doi.org/10.1007/s41468-017-0008-7'
  DOI: 10.1007/s41468-017-0008-7
  page: 241
  type: article-journal
  issued:
    year: 2017
- id: persistence_fisher
  title: 'Persistence fisher kernel: a riemannian manifold kernel for persistence diagrams'
  author:
  - family: Le
    given: Tam
  - family: Yamada
    given: Makoto
  container-title: Neurips
  type: article-journal
  issued:
    year: 2018
- id: distance_calc
  title: 'Geometry Helps to Compare Persistence Diagrams'
  author:
  - family: Kerber
    given: Michael
  - family: Morozov
    given: Dmitriy
  - family: Nigmetov
    given: Arnur
  container-title: ACM Journal of Experimental Algorithmics
  volume: 22
  URL: 'https://doi.org/10.1145/3064175'
  DOI: 10.1145/3064175
  page: 1
  type: article-journal
  issued:
    year: 2017
- id: dependencies
  title: 'Statistical Inference for Persistent Homology applied to fMRI'
  author:
  - family: Abdallah
    given: Hassan
  - family: Regalski
    given: Adam
  - family: Behzad Kang
    given: Mohammad
  - family: Berishaj
    given: Maria
  - family: Nnadi
    given: Nkechi
  - family: Chowdury
    given: Asadur
  - family: Diwadkar
    given: Vaibhav
  - family: Salch
    given: Andrew
  container-title: Github
  URL: 'https://github.com/hassan-abdallah/Statistical_Inference_PH_fMRI/blob/main/Abdallah_et_al_Statistical_Inference_PH_fMRI.pdf'
  type: article-journal
  issued:
    year: 2021
- id: kernel_test
  title: 'A Kernel Statistical Test of Independence'
  author:
  - family: Gretton
    given: Arthur
  - family: Fukumizu
    given: Kenji
  - family: Hui Teo
    given: Choon
  - family: Song
    given: Le
  - family: Scholkopf
    given: Bernhard
  - family: Smola
    given: Alexander
  container-title: NEURIPS
  type: article-journal
  issued:
    year: 2007
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Machine learning and statistical inference of persistence diagrams

Topological data analysis (TDA) is a new area of data science which can find unique, non-linear global structure in whole datasets. The main tool of TDA is called persistent homology, which produces a "shape descriptor" of its input dataset called a "persistence diagram". Two main R packages exist for computing persistence diagrams, however the more flexible of the two, called TDA, stores persistence diagrams in a strange data structure. Methods exist for computing distances and similarities (kernels) between pairs of persistence diagrams, however only distance calculations are available in R packages. Moreover, several papers have used distance and kernel computations in order to perform machine learning or inferences tasks with groups of persistence diagrams, however to date no publically available software in either python or R provides the functionality for these types of analyses. In order to make the power of TDA available to data and machine learning practitioners, a software package must exist which can (1) conver the output of persistent homology into a commonly-used data type for data analyses (a data frame), (2) implement fast versions of both distance and kernel calculations, and (3) provide methods for machine learning and inference using persistence diagrams. It is with these goals that the package TDAML was created.

# Software Review

TDA has gained popularity over the past two decades since the original paper on persistent homology was published, and three main R packages have been created for TDA calculations. The R package TDA is a wrapper for a number of C++ persistent homology packages, including Dionysus, GUDHI and PHAT, and provides distance calculation functionality. The R package TDAstats is a wrapper of the C++ persistent homology engine Ripser, and provides a method for inference of two groups of persistence diagrams (like in @Robinson_Turner), however the distance calculation currently implemented is not a standard distance metric for persistence diagrams. TDAmapper is an R package for the second most-popular TDA method, called mapper, and does not implement any persistent homology calculations. It is worthwhile to note that there are several python packages dedicated to TDA calculations, including scikit-TDA (computing persistence diagrams, distances and kernels between them, and mapper), and giotto-tda (computing persistence diagrams and using persistence diagrams to analyze time series data).

There are a number of current shortcomings of available TDA software. Firstly, in both python and R there is no package which allows for machine learning and inference of persistence diagrams, which greatly limits the types of analyses that can be carried out using standard packages (mainly just visualizing persistence diagrams). In R this is partially because there is no package for kernel calculations of persistence diagrams, but the very slow computation of distances also inhibits the practicality of distance-based inference procedures. Finally, in R the output of persistent homology calculations from the package TDA is a list with an element called "diagram" of class "diagram", and the output of homology calculations from the package TDAstats is a matrix/array. The package TDA seems to be more highly used since it provides some nice additional functionality, including distance calculations and the ability to return the location, in the dataset, of each topological feature found in the persistence diagram. Therefore, the unnatural data type returned by persistent homology calculations in the package TDA may also be limiting the development of TDA applications. 

# Package TDAML

The package TDAML aims to solve the three goals outlined in the introductory paragraph. TDAML is built to use as input the output of persistent homology calculations from the package TDA. Firstly, the function `diagram_to_df` allows the conversion of the output of TDA persistent homology calculations to a data frame. Secondly, the functions `diagram_distance` and `diagram_kernel` allow for fast distance and kernel calculations respetively, and their counterparts `distance_matrix` and `gram_matrix` compute in parallel the (cross) distance and Gram matrices respectively. Thirdly, these distance and kernel calculations are used to perform machine learning and inference on persistence diagrams - methods include dimension reduction with metric multidimensional scaling and kernel princpal components analysis, clustering with kernel k-means, regression and classification with kernel support-vector machines, and inference with distance and kernel calculations looking for group differences and group independence respetively. The kernel machine learning methods implemented are wrappers of the flexible R package for kernel calculations kernlab, with some additional processing steps specific to persistence diagrams. In the subsequent sections we will describe these applications in more detail.

# Background: persistent homology

The main tool of TDA is called persistent homology. TALK ABOUT COOL APPLICATIONS. Persistent homology takes as input either (i) a dataset and a metric for computing distances between points in the dataset, or (ii) a precomputed distance matrix, and outputs a set of 2D points above the diagonal (i.e. where $y > x$) called a persistence diagram. Each point in a persistence diagram represents a topological feature in the dataset of a certain dimension - dimension 0 refers to clusters (connected components), dimension 1 refers to loops (ellipses), dimension 2 refers to voids (spheres) etc. In order to find these topological features in the dataset, a sequence of skeletal approximations to the dataset are constructed which contain topological features, and we track how "significant" each feature is over time. The algorithm proceeds in the following manner: first, if the input is a dataset and distance metric then the distance matrix, storing the distance metric value of each pair of points in the dataset, is computed. Next, a parameter $\epsilon \geq 0$ is grown, and at each $\epsilon$ value we compute a skeletal approximation of the dataset structure by connecting all pairs of points whose distance is at most $\epsilon$, adding a triangle between any triple of points which are all connected, adding a tetrahedron between any quadruple of points which are all connected etc. Note that this process of forming a sequence of skeletal approximations is called a filtration, and other methods exist for forming the approximations however the one described here is the most commonly used, called the Rips-Viertoris complex. Each such skeleton contains some combination of the topological features, i.e. connected components, loops and voids. As $\epsilon$ grows, topological features will be "born" when just enough connections are made, for instance around a loop to make it fully connected, and the $\epsilon$ value at the birth of a topological feature is called the feature's "birth radius". Likewise, once $\epsilon$ becomes very large a topological feature will "die", for instance a loop being fully connected across (destroying the hole that existed at its center), and the $\epsilon$ value at the death of a topological feature is called the feature's "death radius". Therefore, the output persistence diagram has one 2D points for each topological feature found in the filtration, where the $x$-value of the point is the birth radius and the $y$-value is the death radius, hence why every point lies above the diagonal - features die after they are born! Points which are high above the diagonal (measured by $|y-x|$) are said to be "persistent", and likely represent real topological features of the dataset, whereas points near the diagonal likely represent topological noise.

# TDAML methods and underlying theory

## Distance between persistence diagrams

There are several ways to compute distances between persistence diagrams in the same dimension - the most common two are called the 2-wasserstein and bottleneck distances. These techniques find an optimal matching of the 2D points in their input two diagrams, and compute a cost of that optimal matching @distance_calc. A point from one diagram is allowed either to be paired (matched) with a point in the other diagram or its diagonal projection, i.e. the nearest point on the diagonal line $y=x$ using Euclidean distance. Allowing points to be paired with their diagonal projections both allows for matchings of persistence diagrams with different numbers of points (which is almost always the case in practice) and also concretely applies the idea that some points in a persistence diagram really just represent noise. The "cost" value associated with a matching is given by either (i) the maximum of infinity-norm distances between paired points, or (ii) the square-root of the sum of squared Euclidean distances between matched points. The cost of the optimal matching under loss (i) is called the bottleneck distance of persistence diagrams, and the cost of the optimal matching of cost (ii) is called the 2-wasserstein metric of persistence diagrams. Both distance metrics have been used in a number of applications, but the 2-wasserstein metric is able to find more fine-scale differences in persistence diagrams compared to the bottleneck distance. The problem of finding an optimal matching can be solved with the Hungarian algorithm, which is implemented in the R package clue.

Another distance metric between persistence diagrams, which will be useful for kernel calculations, is called the Fisher information metric, $d_{FIM}$ (details can be found in @persistence_fisher). The idea is to represent the two persistence diagrams as probability density functions, with a 2D-Gaussian point mass centered at each point in both diagrams (including the diagonal projections of the points in the opposite diagram), and computing the Fisher information metric between the two distributions. Algorithmically, we calculate the following two vectors: $\rho_1 = [\sum_{u \in D_1 \cup \Delta D_2}N(x;u,\sigma^2)|u \in D_1 \cup \Delta D_2 \cup D_2 \cup \Delta D_1]$ and $\rho_2 = [\sum_{u \in D_2 \cup \Delta D_1}N(x;u,\sigma^2)|u \in D_1 \cup \Delta D_2 \cup D_2 \cup \Delta D_1]$. Then we noramlize them by their sums, $\overline{\rho_1} = \frac{\rho_1}{\sum \rho_1}$ and $\overline{\rho_2} = \frac{\rho_2}{\sum \rho_2}$. Finally, we compute $<\sqrt{\rho_1},\sqrt{\rho_2}>$, i.e. the dot product of the elementwise square root of both of the noramlized vectors.

These distance calculations have been implemented in the `diagram_distance` function, providing a fast native-R computation for all three distance metrics which can be used for standalone analyses or machine learning/inference in other package functions. Note that there is a generalization of the 2-wasserstein distance for any $p \geq 1$, the $p$-wasserstein distance, which can also be computed using the `diagram_distance` function, although analyses where researchers use $p \neq 2$ for distance calculations seem to be rare.

## Metric multidimensional scaling of persistence diagrams

One common dimension-reduction technique in machine learning is called multidimensional scaling (MDS). MDS takes as input an $n$ by $n$ distance (or dissimilarity) matrix $D$, computed from $n$ points in a dataset, and outputs an embedding of those points into a Euclidean space which best preserves the inter-point distances/dissimliarities. To find the optimal embedding we compute $A = -\frac{1}{2}D^2$, and $B$ where $B_{i,j}$ is $A_{i,j}$ subtract the sum the row $i$ and column $j$ means in $A$ plus the overall mean of $A$. Let $v_1,v_2,\dots,v_k$ be the eigenvectors of $B$ ordered from largest to smallest eigenvalue. Then the matrix $V = [v_1|v_2|\dots|v_k]$ is the optimal $k$-dimensional linear embedding of $D$, where the $i$-th row of $V$ is the embedding of the $n$-th point in the dataset \url{https://ncss-wpengine.netdna-ssl.com/wp-content/themes/ncss/pdf/Procedures/NCSS/Multidimensional_Scaling.pdf}. Using the R function `cmdscale` from the package stats we can compute the optimal embedding of a set of persistence diagrams using any of the three distance metrics using the function `diagram_MDS`.

## Testing for group differences of persistence diagrams

Distributions of persistence diagrams can be complicated, and are likely non-parametric. Moreover, there is no unique "mean" of a set of persistence diagrams, so there is no straightforward analogue to a t-test in standard statistics. However, there is an analogue of analysis of variance (ANOVA), for persistence diagrams which can be used to find differences in groups of persistence diagrams. This test was proposed in @Robinson_Turner, and some variations have been suggested. Originally, two groups of persistence diagrams would be compared. The null hypothesis, $H_0$, is that the diagrams from the two groups are generated from the same underlying geometric object. The alternative hypothesis, $H_A$, is that the underlying geometric objects are different. In each dimension a p-value is computed, finding evidence against $H_0$ in those dimensions. A loss function is defined for the two groups which computes the square-root of the sum of within-group sum of squared distances of the persistence diagrams (again we can generalize this measure for other values of $q \geq 1$). We first compute the test statistic as the loss function of the original two groups in each dimension. Then, we permute the group labels many times to obtain shuffled groups and recompute the loss function on these shuffled groups to estimate the null distribution of the test statistic. If the test statistic is smaller than most of these "permutation values" then it is likely that the groups are very different in that dimension, otherwise we cannot say. Formally, a p-value is calculated as $p = \frac{Z+1}{N+1}$ where $Z$ is the number of permutation values smaller than the test statistic, and $N$ is the number of permutations.

This inference procedure is implemented in the `permutation_test` function, with several speedups and additional functionalities. Firstly, the loss function is computed in parallel since distance computations can be expensive. Secondly, we store distance calculations because otherwise we would have to recompute the same distances many times with high likelihood. Additional functionality includes allowing for any number groups (not just two) and to allow for a pairing between groups of the same size as described in @dependencies, which calculates a more realistic estimate of the null distribution of the test statistic when there is a natural pairing between persistence diagrams (like if the groups represent persistence diagrams from the same subject of a study in different conditions for example).

## Kernels between persistence diagrams

## Kernel k-means of persistence diagrams

## Kernel principal components analysis of persistence diagrams

## Kernel support-vector machines of persistence diagrams

## Testing for independence between two groups of paired persistence diagrams



One use of distances between diagrams is embedding a group of diagrams into a low-dimensional space using metric MDS. Metric MDS takes as input a $n$ by $n$ distance matrix and returns an embedding of $n$ points into $d$-dimensional Euclidean space such that the Euclidean distance between the embedded points are as close as possible to the original distance matrix. 

Now that we have defined distances between persistence diagrams we can discuss how to make inferences about groups of persistence diagrams. For the basic idea we will consider two groups of persistence diagrams, $G_{1}$ and $G_{2}$, containing $n_{1}$ and $n_{2}$ persistence diagrams respectively. Each persistence diagram comes from a dataset, which could be viewed as a sample from a true "underlying" manifold (geometric object). Therefore, we might want to test if the groups $G_{1}$ and $G_{2}$ are samples of persistence diagrams coming from the same manifold, $H_{0}$, or if the underlying manifolds are different $H_{A}$. We start by computing a test-statistic, $L$, which is essentially the sum of within-group variances:

$$L = \frac{1}{n_{1}(n_{1}-1)}\sum_{D_{1} \neq D_{2} \in G_{1}}d(D_{1},D_{2})^{q} + \frac{1}{n_{2}(n_{2}-1)}\sum_{D_{1} \neq D_{2} \in G_{2}}d(D_{1},D_{2})^{q}$$

Here, $q$ is some finite exponent which is at least 1, and $d$ is the desired distance of persistence diagrams. $L$ is the test statistic for our hypothesis test.

Since relatively little is known about the distribution of persistence diagrams a permutation test procedure allows us to determine a suitable null distribution for making inferences. We will shuffle the group labels of the persistence diagrams (effectively moving them between the two groups) $N$ times and recompute the test statistic each time to obtain a list of "permutation values" $\{L_{1},\dots,L_{N}\}$. The idea is that if the groups are truly different, i.e. coming from different manifold objects, then there should be relatively small within group variances compared to when we shuffle the group labels. Therefore, let $NZ$ be the number of permutations $i$ such that $L_{i} < L$. Then the p-value of the test is $\frac{NZ+1}{N+1}$.

The extension to multiple groups, noted at the end of the paper @Robinson_Turner, is extremely straightforward and is implemented in this package. The null hypothesis is that all the groups are sampled from the same geometric object, and the alternative hypothesis is that some of the groups are sampled from different geometric objects. The test statistic is still the sum of the (essentially) within-group variances across all the groups, the permutations shuffle the persistence diagrams between all the groups, and the p-value is computed in the same way.

There is one additional functionality that this package provides, and it is an idea that was discussed in @dependencies. One of the assumptions of the inference procedure listed in @Robinson_Turner is that the persistence diagrams are all independent, however this is not always reasonable to assume in practice. For example, suppose we had point cloud data for several subjects in a pharmaceutical study both before and after a drug intervention. We might want to know if the geometric structure of the point clouds changed after the drug intervention, so our two groups of persistence diagrams would be the before intervention and after invention groups. Since there is a pairing between some of the persistence diagrams (the diagrams belonging to the same patient) these diagrams will not be independent of each other. From this perspective, we will only allow permutations which keep the two (or more in the case of many diagram groups) persistence diagrams of each patient (i.e. dependency) separated. So after each permutation each group still contains one persistence diagram per patient. Not only is this more ideal theoretically, but it also offers a practical advantage. If there was truly a difference between the groups then if there are subject differences the permuted-within-group variances would be larger when each group contains one persistence diagram from each subject, enhancing the test power.

In order to use kernel methods for machine learning of persistence diagrams, this package implements the kernel described in @persistence_fisher. The idea is quite simple, let $\sigma,t > 0$. Then the persistence fisher kernel is defined as $\rho(D_1,D_2) = \mbox{exp}(-t*d_{FIM}(D_1,D_2))$, and the Gram matrix of a list of persistence diagrams $D_1,D_2,\dots,D_n$ is the matrix which stores the kernel values: $G[i,j] = \rho(D_i,D_j)$. We can use this kernel to perform several machine learning tasks with persistence diagrams: we can embed a group of persistence diagrams into a low-dimensional space using kPCA by computing eigenvectors of the Gram matrix, we can cluster a group of persistence diagrams using kkmeans, and we can train classification and regression models where the training set is a list of persistence diagrams using kSVM's. TDAML also provides prediction methods for kkmeans, kPCA and kSVM.

To demonstrate the utility of the TDAInference package, we will go through several examples. Firstly, we will give some examples to show that the various distances between persistence diagrams are being computed correctly. Secondly, we will benchmark the distance calculations to show that they are faster than those provided in the TDA package. Thirdly, we will carry out various permutation tests with multiple groups either sampled from the same or different objects to show that the permutation test returns expected results.


```{r setup}
# if(interactive()){
#  devtools::load_all()
# }
library("TDAML")
```

## Case study: `Using the diagram_distance function`

First we will construct a pair of small diagrams for which we can easily calculate by hand what their distance should be (according to the various types of distance metrics available in this package), and then we will do the same with diagrams generated from the TDA package. Let $D_1 = \{(2,3)\}$ be the diagram with a single point $(2,3)$, and let $D_2 = \{(2,3.1),(5,6)\}$ have two points. We will construct these diagrams as data frames, as if they were the output of the diagram_to_df function, and plot them:

```{r}
# D1
D1 = data.frame(dimension = 0,birth = 2,death = 3)
plot(x = D1$birth,y = D1$death,xlab = "Birth",ylab = "Death",main = bquote("" ~ D[1]),xlim = c(0,6),ylim = c(0,6),pch = 16)
abline(a = 0,b = 1)

# D2
D2 = data.frame(dimension = 0,birth = c(2,5),death = c(3.1,6))
plot(x = D2$birth,y = D2$death,xlab = "Birth",ylab = "Death",main = bquote("" ~ D[2]),xlim = c(0,6),ylim = c(0,6),pch = 16)
abline(a = 0,b = 1)

```

Now, let's compute the distance between $D_1$ and $D_2$ using the various distance metrics, and confirm that  it's clear what the distance value should be. Note that for all metrics the optimal matching is the same: $(2,3)$ is matched with $(2,3.1)$ and $(5,6)$ is matched with its diagonal projection, $(5.5,5.5)$. Therefore, using the 2-wasserstein metric (like in the package TDA) the distance should be $\sqrt{0.1^2+0.5^2} \approx 0.509902$:

```{r}
diagram_distance(D1 = D1,D2 = D2,p = 2,distance = "wasserstein",dim = 0)
```

The same distance but with $p = 3$ should be $(0.1^3+0.5^3)^{1/3} \approx 0.5013298$:

```{r}
diagram_distance(D1 = D1,D2 = D2,p = 3,distance = "wasserstein",dim = 0)
```

The persistence Fisher distance between $D_1$ and $D_2$ using sigma = 1 should be calculated as follows. $\rho_1 = \frac{1}{2\pi}(e^0+e^{-0.2525}+e^{-9.25},e^{-0.2525}+e^0+e^{-8.7025},e^{-9.25}+e^{-8.7025}+e^0,e^{-0.005}+e^{-0.3025}+e^{-9.005},e^{-9}+e^{-8.9525}+e^{-0.25},e^{-0.25}+e^{-0.0025}+e^{-9})$ and $\rho_2 = \frac{1}{2\pi}(e^{-0.005}+e^{-9}+e^{-0.25},e^{-0.3025}+e^{-8.9525}+e^{-0.0025},e^{-9.005}+e^{-0.25}+e^{-9},e^{0}+e^{-8.705}+e^{-0.305},e^{-8.705}+e^{0}+e^{-9.25},e^{-0.305} + e^{-9.25} + e^{0})$. Then the distance metric should equal $<\sqrt{\overline{\rho_1}},\sqrt{\overline{\rho_2}}> \approx 0.9984164$. When we use the diagram distance function with distance set to fisher we see that we get the same result:

```{r}
diagram_distance(D1 = D1,D2 = D2,distance = "fisher",dim = 0,sigma = 1)
```

Finally, we can compute the standard bottleneck distance between the diagrams, which is just the maximum bottleneck distance between any pair of matched points, 0.5:

```{r}
diagram_distance(D1 = D1,D2 = D2,p = Inf,distance = "wasserstein",dim = 0)
```

We include more tests of correctness for the diagram_distance function in the test folder, should the reader be interested in further validation.

<!-- OLD -->

<!-- $D_3 = \{(1,1.1),(3,3.1)\}$. We can visualize the three diagrams below: -->

<!-- ```{r} -->
<!-- D3 = data.frame(x = c(1,3),y = c(1.1,3.1)) -->
<!-- ``` -->

<!-- We can do similar calculations for each of the other pairs $(D_1,D_3)$ and $(D_2,D_3)$. For $(D_1,D_3)$ the optimal matchine is each point with its diagonal projection: $(2,3)$ with $(2.5,2.5)$, $(1.1.1)$ with $(1.05,1.05)$ and $(3,3.1)$ with $(3.05,3.05)$. It is easy to verify that the distance should be $\sqrt{0.05^2+0.05^2+0.5^2} \approx 0.5049752$ for the 2-wasserstein metric, $\sqrt{0.05^2+0.05^2+0.05^2+0.05^2+0.5^2+0.5^2} \approx 0.7141428$ for the $p = 2$ Turner metric, and $0.5$ for the Infinity norm metric. We see that those calculations were carried out correctly by the digram_distance function as well: -->

<!-- It is easy to verify that the distances between $D_2$ and $D_3$ are computed properly, but it covers similar cases to previous distances so we will not write out the verification in this document. -->

## Case study: `Benchmarking the diagram_distance function`

Now we will benchmark the runtime of the distance calculations on pairs of different-sized random persistence diagrams compared to calculations using the package TDA. We will first simulate random diagrams of various sizes to compare runtimes, with $n$ points ($n \in \{100,200,300,\dots,10000\}$) with 10 iterations at each value of $n$. Each diagram has birth values of 0.1 multiplied by a random sample of a chi squared distribution with 10 degrees of freedom, and the death was the birth plus 0.1 times a sample from a chi squared distribution with 1 degree of freedom. The code took much too long to generate with this vignette, but can be viewed in the .Rmd file for this vignette. The runtimes were computed on a Windows 10 64-bit machine, an Intel(R) Core(TM) i7-7700 CPU @ 3.60GHz 3.60 GHz processor with 8 cores and 64GB of RAM. The benchmarking plot can be seen below:

```{r,echo = F,warning=F,out.width="80%",out.height="90%"}
# simulate random barcodes with sizes 100,200,...,1000
# library(TDA)
# runtimes_random <- data.frame(n_row = numeric(),package = character(),time_in_sec = numeric())
# for(n_row in seq(100,1000,100)){
# 
#   for(iteration in 1:10)
#   {
# 
#     # simulate pair of diagrams, points mainly close to the diagonal
#     diagram1 = data.frame(dimension = rep(0,n_row),birth = 0.1*rchisq(n = n_row,df = 10))
#     diagram1$death = diagram1$birth + 0.1*rchisq(n = n_row,df = 1)
#     diagram2 = data.frame(dimension = rep(0,n_row),birth = 0.1*rchisq(n = n_row,df = 10))
#     diagram2$death = diagram2$birth + 0.1*rchisq(n = n_row,df = 1)
# 
#     # benchmark time, in seconds, for both package's distance calculations (2-wasserstein distance in TDAInference)
#     start_time_TDAInference = Sys.time()
#     d_TDAInference = diagram_distance(D1 = diagram1,D2 = diagram2,dim = 0,p = 2,distance = "wasserstein")
#     end_time_TDAInference = Sys.time()
#     time_diff_TDAInference = as.numeric(end_time_TDAInference - start_time_TDAInference,units = "secs")
#     diagram1 = list(diagram = diagram1)
#     diagram2 = list(diagram = diagram2)
#     diagram1$diagram = as.matrix(diagram1$diagram)
#     diagram2$diagram = as.matrix(diagram2$diagram)
#     class(diagram1$diagram) = "diagram"
#     class(diagram2$diagram) = "diagram"
#     start_time_TDA = Sys.time()
#     d_TDA = TDA::wasserstein(Diag1 = diagram1$diagram,Diag2 = diagram2$diagram,dimension = 0,p = 2)
#     end_time_TDA = Sys.time()
#     time_diff_TDA = as.numeric(end_time_TDA - start_time_TDA,units = "secs")
# 
#     runtimes_random = rbind(runtimes_random,data.frame(n_row = n_row,package = "TDAInference",time_in_sec = time_diff_TDAInference))
#     runtimes_random = rbind(runtimes_random,data.frame(n_row = n_row,package = "TDA",time_in_sec = time_diff_TDA))
# 
#   }
# 
# }
# 
# # compute means and sd's at each value of rows for both packages
# summary_table = data.frame(n_row = numeric(),mean = numeric(),sd = numeric(),package = character())
# for(n_row in seq(100,1000,100))
# {
#   
#   for(p in c("TDAInference","TDA"))
#   {
#     
#     summary_table = rbind(summary_table,data.frame(n_row = n_row,mean = mean(runtimes_random[which(runtimes_random$n_row == n_row & runtimes_random$package == p),3]),sd = sd(runtimes_random[which(runtimes_random$n_row == n_row & runtimes_random$package == p),3]),package = p))
#     
#   }
#   
# }

# since this code takes a very long time to run, we will hard code the table that resulted from the above section^
summary_table = data.frame(n_row = rep(seq(100,1000,100),each = 2),mean = c(0.087,0.137,0.278,1.880,0.859,8.742,1.782,28.908,3.928,63.544,6.648,141.373,11.158,252.689,18.409,427.515,28.220,725.520,39.675,1024.007),sd = c(0.021,0.010,0.018,0.170,0.100,0.630,0.161,1.849,0.245,3.927,0.455,12.202,0.806,14.315,1.350,37.318,2.675,63.963,2.983,102.328),package = rep(c("TDAInference","TDA"),10))

# plot table
plot(summary_table$n_row[summary_table$package=="TDA"], summary_table$mean[summary_table$package=="TDA"], type="b",
     xlim=range(summary_table$n_row), ylim=range(0,summary_table$mean+1.96*summary_table$sd),xlab = "Points in diagram",ylab = "Mean execution time (sec)",main = "Benchmarking diagram_distance function")
lines(summary_table$n_row[summary_table$package=="TDAInference"], summary_table$mean[summary_table$package=="TDAInference"], col=2, type="b")
legend(x = 200,y = 800,legend = c("TDAInference","TDA"),col = c("red","black"),lty = c(1,1),cex = 0.8)
arrows(summary_table$n_row, summary_table$mean-1.96*summary_table$sd/sqrt(10), summary_table$n_row, summary_table$mean+1.96*summary_table$sd/sqrt(10), length=0.05, angle=90, code=3)

```

The TDAInference diagram\_distance function gives a significant speedup compared to the TDA wasserstein function (with similar results expected for the bottleneck distance function since the two use the same underlying engine). As the diagrams get larger the speedup appears to increase, with almost a 26x speedup for diagrams with 1000 points, however this ratio looks like it would plateau around 30x above 1000 points, or at least it would grow slowly. 

Now, we will repeat the same experiment but instead computing diagrams from shapes provided in the TDA package (a uniformly sampled Torus and sphere) with the same number of data points as in the previous example, $n \in \{100,200,\dots,1000\}$ with 10 repetitions per size. The homology calculations had a maximum dimension of two, and the maximum scale for each was 2 (for the Torus) and 1 (for the sphere). The Torus had tube radius 1 and cross-radius 2. Again, to avoid a long build for this vignette the results were computed on the same computer as listed above, and can be generated using the code in the .Rmd file. The plot is as follows:

```{r,echo = F,warning=F}
# generate persistence diagrams from circles, Tori and spheres with  100,200,...,1000 data points.
# library(TDA)
# runtimes_shape <- data.frame(n_row = numeric(),package = character(),time_in_sec = numeric())
# for(n_row in seq(100,1000,100)){
# 
#   for(iteration in 1:10)
#   {
# 
#     # simulate pair of diagrams from the desired shapes
#     diagram_torus = ripsDiag(X = TDA::torusUnif(n = n_row,a = 1,c = 2),maxdimension = 2,maxscale = 2)
#     diagram_sphere = ripsDiag(X = TDA::sphereUnif(n = n_row,d = 2,r = 1),maxdimension = 2,maxscale = 1)
#     
#     # compute their wasserstein distances in all dimensions and benchmark
#     start_time_TDAInference = Sys.time()
#     diagram_distance(D1 = diagram_torus,D2 = diagram_sphere,dim = 0,p = 2,distance = "wasserstein")
#     diagram_distance(D1 = diagram_torus,D2 = diagram_sphere,dim = 1,p = 2,distance = "wasserstein")
#     diagram_distance(D1 = diagram_torus,D2 = diagram_sphere,dim = 2,p = 2,distance = "wasserstein")
#     end_time_TDAInference = Sys.time()
#     time_diff_TDAInference = as.numeric(end_time_TDAInference - start_time_TDAInference,units = "secs")
#     
#     start_time_TDA = Sys.time()
#     TDA::wasserstein(Diag1 = diagram_torus$diagram,Diag2 = diagram_sphere$diagram,dimension = 0,p = 2)
#     TDA::wasserstein(Diag1 = diagram_torus$diagram,Diag2 = diagram_sphere$diagram,dimension = 1,p = 2)
#     TDA::wasserstein(Diag1 = diagram_torus$diagram,Diag2 = diagram_sphere$diagram,dimension = 2,p = 2)
#     end_time_TDA = Sys.time()
#     time_diff_TDA = as.numeric(end_time_TDA - start_time_TDA,units = "secs")
# 
#     runtimes_shape = rbind(runtimes_shape,data.frame(n_row = n_row,package = "TDAInference",time_in_sec = time_diff_TDAInference))
#     runtimes_shape = rbind(runtimes_shape,data.frame(n_row = n_row,package = "TDA",time_in_sec = time_diff_TDA))
# 
#   }
#   print(paste0("Done ",n_row," rows"))
# 
# }
# 
# # compute means and sd's at each value of rows for both packages
# summary_table = data.frame(n_row = numeric(),mean = numeric(),sd = numeric(),package = character())
# for(n_row in seq(100,1000,100))
# {
# 
#   for(p in c("TDAInference","TDA"))
#   {
# 
#     summary_table = rbind(summary_table,data.frame(n_row = n_row,mean = mean(runtimes_shape[which(runtimes_shape$n_row == n_row & runtimes_shape$package == p),3]),sd = sd(runtimes_shape[which(runtimes_shape$n_row == n_row & runtimes_shape$package == p),3]),package = p))
# 
#   }
# 
# }

# here is the output that I got on my local computer
summary_table = data.frame(n_row = rep(seq(100,1000,100),each = 2),mean = c(0.133,0.220,0.398,3.430,0.953,17.054,1.955,53.896,3.760,142.524,6.604,294.552,10.466,601.802,16.169,1130.289,23.458,2091.953,35.098,3518.517),sd = c(0.022,0.016,0.035,0.186,0.094,1.1889,0.130,4.501,0.211,5.924,0.772,11.244,0.623,9.064,0.989,43.385,1.989,88.432,2.747,172.684),package = rep(c("TDAInference","TDA"),10))

# plot table
plot(summary_table$n_row[summary_table$package=="TDA"], summary_table$mean[summary_table$package=="TDA"], type="b",
     xlim=range(summary_table$n_row), ylim=range(0,summary_table$mean+1.96*summary_table$sd),xlab = "Points in shapes",ylab = "Mean execution time (sec)",main = "Benchmarking diagram_distance function on Torus and Sphere")
lines(summary_table$n_row[summary_table$package=="TDAInference"], summary_table$mean[summary_table$package=="TDAInference"], col=2, type="b")
legend(x = 200,y = 2000,legend = c("TDAInference","TDA"),col = c("red","black"),lty = c(1,1),cex = 0.8)
arrows(summary_table$n_row, summary_table$mean-1.96*summary_table$sd/sqrt(10), summary_table$n_row, summary_table$mean+1.96*summary_table$sd/sqrt(10), length=0.05, angle=90, code=3)

```

Once again, the runtime of the diagram\_distance function is significantly smaller. The ratio of runtimes grew at each subsequent number of tested data points in the shapes, with a 100x speed up with 1000 data points. These two simulations point towards TDAInference a fast distance calculation between persistence diagrams, and that performing many such computations, like in the case of a permutation test of persistence diagrams, could become much more feasible.

## Case study: `Using the permutation_test function`

The permutation\_test function can be used to make statistical inferences about any number of groups of persistence diagrams computed from the TDA package. To see this functionality, we will create three groups of diagrams with unequal numbers of elements:

```{r,echo = T}

g1 <- lapply(X = 1:3,FUN = function(X){

  diag <- TDA::ripsDiag(TDA::circleUnif(n = 100,r = 1),
                                   maxscale = 1,
                                   maxdimension = 1)
  df <- diagram_to_df(d = diag)
  return(df)

})

g2 <- lapply(X = 1:3,FUN = function(X){

  diag <- TDA::ripsDiag(TDA::torusUnif(n = 100,c = 2,a = 1),
                                   maxscale = 1,
                                   maxdimension = 1)
  df <- diagram_to_df(d = diag)
  return(df)

})

g3 <- lapply(X = 1:2,FUN = function(X){

  diag <- TDA::ripsDiag(TDA::sphereUnif(n = 100,r = 1,d = 2),
                                   maxscale = 1,
                                   maxdimension = 1)
  df <- diagram_to_df(d = diag)
  return(df)

})

```

Since the groups are not of the same size, we cannot set the paired argument to be True. The permutation test could be carried out, with mostly default parameters, by

```{r,warning=FALSE}
perm_test = permutation_test(g1,g2,g3,verbose = TRUE)
```

We can access the various components of the test easily. The p-values in each dimension can be viewed with

```{r}
perm_test$p_value
```

The test statistic can be viewed with

```{r}
perm_test$test_statistic
```

The permutation values can be accessed in a list, with one element for each desired dimension.

```{r}
perm_test$permvals$`0`
```

Other parameter values that can be viewed are the dimensions in which the test was carried out and the runtime of the test. Lastly, we will carry out the same calculation, but with pairing. To that end all groups must have the same number of elements:

```{r}
g3[3] <- g3[1]
perm_test_paired = permutation_test(g1,g2,g3,verbose = T,paired = T)
```

This paired parameter should be set to True when there is a natural pairing between the diagrams in the groups. For example, if each group represents the persistence diagram of (in this case) three subject's fMRI data, and the three groups represent three different fMRI tasks, then a pairing between the groups exists. Setting the pairing parameter to True in this case can reduce the variance in permutation values, thereby reducing the likelihood of spurious significant test results.


## References


